Pool Model - A unified Mathematical\\ Modelling Framework for Food Science

### 2025-11-04 paper/pool_model.tex
fix some formulations/typos

        Simulation snapshots at $1.04$\unit{\day},$2.08$\unit{\day},$3.13$\unit{\day} and $4.17$\unit{\day} corresponding to the parameters given in table~table~\ref{tab:spatial_limit_init}.

@@

We see that in the highly-diffusive homogenous case of subfigure (\textbf{A}), the pool model is a good fit to the \ac{abm} model.
Heterogeneity and slower diffusion cause larger differences between the two models as can be seen in subfigures (\textbf{B}) and \textbf{C}.

@@

In subfigure \textbf{C}, due to the heterogeneous cell distribution (see Appendix~\ref{fig:supplement-spatial-initial-config-cells}), overall inhibition of species $B$ is significantly reduced.
Subfigure (\textbf{D}) shows how a randomized homogenous initial distribution can lead to even more notable effects of inhibition.

@@

The presented pool model was used to describe the dynamics of one- and two-species systems containing the main growth stages and interactions between bacteria in the form of cooperation and competition.
Our approach produces results which exhibit similarities to other well-known predictive microbiology models such as the Baranyi-Roberts model while providing better explainability.
The inclusion of additional metabolic processes or larger number of cell-species can be achieved by extending the list of possible states and defining transitions between them.
As an example, the description of the quorum sensing phenomena can be modelled by adding the equation for the released autoinducer molecules as a part of the microenvironment.
These molecules then affect the same-species bacteria growth controlling the size of the population.
The described above approach provides a primary predictive model with transition rates between different pools as parameters.
To include the influence of external factors, i.e., temperature, pH, pressure, etc., one can simply use the secondary models for the transition rates, e.g.,, the Ratkowsky model for temperature dependence~\citep{ratkowsky_relationship_1982}.\\
By constructing an \ac{abm}, we have exemplified how close the devised model is to the individual biological 
processes that occur within the bacteria.
This mechanistic single-cell level of understanding of our \ac{abm} is a perspective which can not be easily replicated by other spatial models such as \acp{pde}.
We have investigated the impact of spatial effects in the form of varying diffusion parameters and different initial spatial distributions of bacteria.
Our findings reinforce that the \ac{ode} model is a good simplification in the highly-diffusive regime within a mostly homogenous spatial bacterial distribution.
However, deviations to these assumptions can lead to non-negligible effects, influencing the growth of the colony.

@@

We introduced the pool model as a framework to unify existing approaches and allows researchers to follow a mechanistic approach during model development.
It is able to capture the effects of many established models and provides a lens which is useful to generalize existing systems and results in better overall explainability.
By modeling effects such as mutual inhibition on a single-cell level within an \ac{abm}, we have exemplified that the pool model directly builds on these more fundamental concepts.
It can thus aid in the mechanistic understanding of the underlying processes, allowing researchers to deepen their knowledge about microbial growth within food science.
Going forward, this framework can provide a starting point for a more abstract mathematical formulation which should be able to cover a much wider class of systems.

>>>>

        Simulation snapshots at $1.04$\unit{\day},$2.08$\unit{\day},$3.13$\unit{\day} and $4.17$\unit{\day} corresponding to the parameters given in Table~\ref{tab:spatial_limit_init}.

@@

We see that in the highly-diffusive homogenous case of subfigure (A), the pool model is a good fit to the \ac{abm} model.
Heterogeneity and slower diffusion cause larger differences between the two models as can be seen in subfigures (B) and (C).

@@

In subfigure (C), due to the heterogeneous cell distribution (see Appendix~\ref{fig:supplement-spatial-initial-config-cells}), overall inhibition of species $B$ is significantly reduced.
Subfigure (D) shows how a randomized homogenous initial distribution can lead to even more notable effects of inhibition.

@@

The presented pool model was used to describe the dynamics of one- and two-species systems containing the main growth stages and interactions between bacteria in the form of cooperation and competition.
Our approach produces results which exhibit similarities to other well-known predictive microbiology models such as the Baranyi-Roberts model while providing better explainability.
The inclusion of additional metabolic processes or larger number of cell-species can be achieved by extending the list of possible states and defining transitions between them.
As an example, the description of the quorum sensing phenomena can be modelled by adding the equation for the released autoinducer molecules as a part of the microenvironment.
These molecules then affect the same-species bacteria growth controlling the size of the population.
The approach described above provides a primary predictive model with transition rates between different pools as parameters.
To include the influence of external factors, i.e., temperature, pH, pressure, etc., one can simply use the secondary models for the transition rates, e.g., the Ratkowsky model for temperature dependence~\citep{ratkowsky_relationship_1982}.\\
By constructing an \ac{abm}, we have exemplified how close the devised model is to the individual biological 
processes that occur within the bacteria.
This mechanistic single-cell level of understanding of our \ac{abm} is a perspective which can not be easily replicated by other spatial models such as \acp{pde}.
We have investigated the impact of spatial effects in the form of varying diffusion parameters and different initial spatial distributions of bacteria.
Our findings reinforce that the \ac{ode} model is a good simplification in the highly-diffusive regime within a mostly homogenous spatial bacterial distribution.
However, deviations to these assumptions can lead to non-negligible effects, influencing the growth of the colony.

@@

We introduced the pool model as a framework to unify existing approaches and allow researchers to follow a mechanistic approach during model development.
It is able to capture the effects of many established models and provides a lens which is useful to generalize existing systems and results in better overall explainability.
By modeling effects such as mutual inhibition on a single-cell level within an \ac{abm}, we have exemplified that the pool model directly builds on these more fundamental concepts.
It can thus aid in the mechanistic understanding of the underlying processes, allowing researchers to deepen their knowledge about microbial growth within food science.
Going forward, this framework can provide a starting point for a more abstract mathematical formulation which should be able to cover a much wider class of systems.
